---
title: "Exploratory"
author: "Kushin"
date: "10/20/2018"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(ggplot2)
library(dplyr)
raw_data<- read.csv('itineraries-2017.csv')

#Subsetting for flights that leave from NY state
ny_data<- raw_data%>%filter(ORIGIN_STATE_ABR=='NY')
#Linear model where the predictor is a categorical variable for the the airport from which the plane departs
ny_model<- lm(MARKET_FARE~factor(ORIGIN), data= ny_data)
#summary(ny_model)

#creating a df which contains average fare and layovers for all flights leaving from NY, organized by state
mean_nyo_data<-raw_data%>%filter(ORIGIN_STATE_ABR=='NY')%>%group_by(DEST_STATE_ABR)%>%summarise(mean_fare= mean(MARKET_FARE), mean_dist = mean(MARKET_DISTANCE), num_lay= mean(numLayovers))

#creating a df which contains average fare and layovers for all flights arriving at NY, organized by state
mean_nyi_data<- raw_data%>%filter(DEST_STATE_ABR=='NY')%>%group_by(ORIGIN_STATE_ABR)%>%summarise(mean_fare= mean(MARKET_FARE), mean_dist = mean(MARKET_DISTANCE), num_lay= mean(numLayovers))


#Full and reduced models for obtaining expected mean fare from mean distance and mean layovers. Reduced model does not have the interaction term. This set of models is for outgoing flights data
model1<-lm(mean_fare~ mean_dist+num_lay+num_lay*mean_dist, data= mean_nyo_data)
model2<-lm(mean_fare~ mean_dist+num_lay, data= mean_nyo_data)
#summary(model1)
#summary(model2)

#F-test to compare full and reduced models
ft1<-anova(model1,model2)
#ft1


#Full and reduced models for obtaining expected mean fare from mean distance and mean layovers. Reduced model does not have the interaction term. This set of models is for incoming flights data
model3<-lm(mean_fare~ mean_dist+num_lay+num_lay*mean_dist, data= mean_nyi_data)
model4<-lm(mean_fare~ mean_dist+num_lay, data= mean_nyi_data)
summary(model3)
#summary(model4)

#F-test to compare full and reduced models
ft2<-anova(model3,model4)
#ft2

#Exploratory plot to look at how distance and number of layovers relates to fares for outgoing flights

ggplot()+geom_point(aes(x=mean_dist, y=mean_fare, col=num_lay), data=mean_nyo_data)
ggplot()+geom_point(aes(x=mean_dist, y=mean_fare, col=num_lay), data=mean_nyi_data)

#Generic code for obtaining conf intervals for slope estimates
#confint(model3, 'mean_dist:num_lay' ,level=0.95)



```


**Intro**  
Given that for the data we have, each of our observations relate to the state of NY in some way, we thought it would be interesting to look at airfare trends as they relate to flights flying into and leaving the state, and check to see if there are any notable differences between these two groups of flights. We were also interested in observing any overall trends in domestic U.S. flights, so as to provide potential passengers with knowledge that they may not be aware of when booking tickets at any given point in time between a particular set of states.

Some commonplace observations that frequent flyers make is that for flights across larger distances, breaking the journey into smaller legs helps reduce the overall cost of the journey. Thus, when flying across large distances on a budget, flyers sometimes look out for flights with one or more layovers. Perhaps the cost in time is sometimes worth the discount in ticket fare.
Another observation, which is almost patent, is that flights become much more expensive during certain times of year - notably during the holiday season. So we might expect journeys during a particular time of year to be more expensive than the same journey during other times.
With these motivating questions, we decided to conduct some statistical analyses to see how well these folk observations hold up to empirical data.

**Methods**  

Since we were interested in looking at differences between journeys that were ending in NY versus those that were beginning at NY, we divided our dataset into itineraries of flights that departed from NY and flights that landed in NY.
There were 4 journeys in the dataset that both began and ended in NY, and we included them in both groups.
Henceforth, the group that included flights that landed in NY will be referred to as the incoming group and the group that included flights that departed from NY will be referred to as the outgoing group.

For each group we found the average flight fare and average number of layovers for flights between NY and all other states. For the incoming group there was data present for 40 states whereas for the outgoing group there was data for 37 states. Since there are multiple airports per state and we were looking at state-state itineraries, we also decided to find the average distance between the starting and ending points of the journey for a given pair of states.

Since flight fare was our dependent variable of interest, we began by fitting 2 multiple linear models, one for each group, where the predictors were mean distance of the journey and mean number of layovers.
Our model can be represented by the following equation - 
$$\widehat{Fare} = \beta_0 +\beta_1 Mean Distance +\beta_2MeanLayovers $$
We also fit a version of these models with an interaction term for mean distance and mean number of layovers. The rationale for this was that while we expect longer journeys to naturally be more expensive, having a journey with one or more layovers may mitigate some of those costs.
This model could be expressed as - 

$$\widehat{Fare} = \beta_0 +\beta_1 Mean Distance +\beta_2MeanLayovers + \beta_3MeanDistance\times MeanLayovers$$
**Results**

Incoming Group -  

For our 'reduced' model without the interaction term, we achieve statistical significance for the estimated slope coefficient for mean distance. The estimated slope coefficient for mean distance is 0.16776 (95 CI: 0.1043711 - 0.2311455). This model has a Multiple R-squared value of 0.5465.

For our 'full' model, which includes the interaction term, we achieve statistical significance for the slope coefficients for the main effect of average layovers and the interaction term as well. The estimated slope coefficient for average number of layovers is -238.57968 (95 CI: -448.8797 - -28.27968) and the estimated slope coefficient for the interaction term is 0.20031 (95 CI: 0.08791061 - 0.3127084 ).
The Multiple R-squared for this model is 0.6672

An F-test comparing the full and reduced models yielded an F value of 9.0017 (p = 0.005103)

Outgoing Group - 

For our 'reduced' model without the interaction term, we achieve statistical signficance for the slope coefficients for mean distance and mean number of layovers. The estimated slope coefficient for mean distance is 0.06566 (95 CI: 0.02868519 - 0.1026327) and for mean layovers is 139.15076 (95 CI: 56.69137 - 221.6102). The Multiple R-squared value for this model is 0.5916. 

For our 'full' model, which includes the interation term, we fail to achieve statistical significance for the estimated slope coefficients for any of the main terms, but we do achieve statistical significance for the estimate slope coefficient for the interaction term. The estimated slope coefficient for the interaction term is 7.076e-02 (95 CI: 0.02277662 - 0.11874). The Multiple R-squared value for this model is 0.6791.

An F-test comparing the full and reduced models yielded an F value of 13.063 (p = 0.0009133)


**Discussion**

From our results, we can see that for journeys that terminate in NY, mean distance between the starting and ending points of the journey positively influence expected fare, that is higher the mean distance higher is the expected fare. But this mean distance is only a significant predictor of fare when there is no interaction term included for mean distance and mean layovers. When this term is included we see that higher the number of layovers, lower is the expected price and that there is a positive interaction between the terms. 
We see that the full model has a higher Multiple R-squared value, and an f-test between these two models reveals that there is a siginificant effect of adding the interaction term. So for incoming flights, our assumptions about layovers reducing costs of flights hold true with the effect being strengthened by larger distances.


Now for journeys that begin in NY, when we don't include an interaction term we see that that both mean distance and number of layovers significantly affect the expected fare in a positive direction. While this makes sense for mean distance it goes against our assumptions that higher number of layovers would result in higher fares.
When we include the interaction term for mean distance and mean layovers we see the significant effect of these main predictors vanish, and we get a significant interaction term. What this might indicate is that while neither of the main predictors alone help estimate fare in our dataset, when both are present the higher their value the higher the expected price. This again goes against our intuitions for number of layovers. And as before the full model does have a higher Multiple R-squared value indicating that it explains more variability in our dataset. And an f-test does reveal that there is a significant effect of adding the interaction term to the reduced model.


From our full models we see that distance alone isn't always a reliable predictor of how expensive a journey will be, regardless of whether it ends or begins in NY. For journeys that end in NY there seems to be a cleaner picture regarding the effect of including more layovers to reduce the cost of the journey. For the case of journeys that begin in NY, the effect of layovers is not as transparent, and seems to work in a direction opposite to what we might expect. These could be because we're aggregating over airports, or because of the limited nature of this dataset. 

```{r message=FALSE, warning=FALSE, echo =FALSE, paged.print=FALSE}
confint(model3, 'mean_dist:num_lay' ,level=0.95)

model3<- lm(MARKET_FARE~MARKET_DISTANCE+numLayovers+QUARTER+numLayovers*MARKET_DISTANCE,data=raw_data)
summary(model3)

for(i in 1:nrow(raw_data)){
    if(as.character(raw_data$ORIGIN_STATE_ABR[i])==as.character(raw_data$DEST_STATE_ABR[i])){
      print(i)
    }

}


ggplot() + geom_point(aes(x=MARKET_DISTANCE,y=MARKET_FARE, col=QUARTER),data=raw_data)+ylim(0,2000)
ggplot() + geom_point(aes(x=MARKET_DISTANCE,y=MARKET_FARE, col=numLayovers),data=raw_data)+ylim(0,2000)
ggplot() + geom_boxplot(aes(x=factor(QUARTER), y = MARKET_FARE),data=raw_data)
```